Grade Retention and Mixed Race

### 2022-08-30 products/paper/main.qmd
Rework first section of Where do Biracials fit
This is the part focused on "pure" race explanations for the hierarchy and the classification of biracials

>>>>

To better understand where biracial individuals might fit, it is necessary to understand what is producing a contemporary system of racial inequality overall. Here, I distinguish three mechanisms that may help us understand the biracial position within this racial hierarchy. First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level. Second, I consider arguments regarding different distributions of resources across racial groups that produce different average arguments. Thirs, I consider arguments regarding how the nature of hybridity itself may affect outcomes.

Discrimination and bias by race play an important role in overall levels of racial inequality. Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unkwnowningly rely upon and reinforce in social interactions.  Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.

Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into institutional systems. Such systemic racism has two effects. First, it provides norms and ideologies that create, support, and reinforce the cognitive schemas that individuals use in everyday practice. Second, norms, laws, and organizational practices can have built in biases that channel rewards to certain racial groups. In our current "colorblind" regime, such systems may be formally race neutral but produce disparate impacts by race.

The experiences of biracial individuals within such a system of discrimination depend almost entirely on how they are classified by others. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups than the other, they will receive the same treatment and reap similar rewards or suffer similar penalties.

The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concering Black ancestry. According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries. Davis (==CITE==) generalizes this practice to the idea of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group. A symmetric, but opposite, pattern of *hyperdescent* has governed the experiences of individuals with American Indian ancestry (==CITE==).

Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the usefulness of this paradigm to both the current and historical situations. Although often couched in generalizable term, hypodescent and hyperdescent are actually only observed in two specific historical cases involving Black ancestry and American Indian ancestry. As CITE note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification. In this framework, among the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.

The development of these particular norms may also reflect the historical regime in which they were developed (CITE). In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us completely to understand the experiences of racialized groups such as Latinos and Asians. Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era (CITE).

If outcomes for biracial individuals all closer to one of their constituent groups, then we may have stronger evidence about such norms of classification. However, what if those outcomes fall in-between the two groups? How might classification and discrimination produce such an in-between status? There are several possibilities here.

First, racial ambiguity in appearance may lead to more inconsistent classification into single race categories across observers. Even if the norms consistently indicate a certain type of classification, correct identification by observers is still a probabilistic phenomenon in all social interactions and likely to be more heterogenous for biracial individuals. As a result, biracial individuals will experience a lower sum total of discrimination and bias affecting the lower status group but will also not reap the same rewards as a member of the higher status group.

Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination. In other words, the experience of discrimination may be "softened" because their ambigious physical appearance provokes less racial antipathy. The US in fact already has a long history of this process. Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes. Thus light skin can moderate (but not eliminate) the stigma of Blackness even among monoracial individuals.

Third, it is also possible that biracial individuals may be classified into a distinct "middleman" minority position. This kind of middleman status is common in Latin America (CITATIONS) and Bonilla-Silva (CITE) has argued that this may be a possible future for the US as well. In this case, many biracial groups may occupy a nebulous and ill-defined middleman or buffer class position in-between whites and darker-skinned non-whites, helping to cement a new kind of racial hierarchy.
